\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-07-15 tex/paper.tex
save

We assume that the stochastic generative process that defines $p(\mathbf{x}|\bftheta)$ is
specified through a non-differentiable deterministic function $g(\cdot; \bftheta) : {\cal Z} \to
\mathbb{R}^d$. Operationally, %That is, we consider
\begin{equation}\label{eqn:p_theta}
    \mathbf{x} \sim p(\mathbf{x}|\bftheta) \equiv \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta)
\end{equation}
such that the density $p(\mathbf{x}|\bftheta)$ can be
written as
%considered as a non-differentiable reparameterization of the random variable $\bfz$
%\begin{equation}\label{eqn:p_x_sim}
%    p(\mathbf{x}|\bftheta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\bfz:g_i(\bfz;\bftheta) \leq x_i\}} p(\bfz|\bftheta) \mu(d\bfz),
%\end{equation}
\begin{equation}\label{eqn:p_x_sim}
    p(\mathbf{x}|\bftheta) = \int_{\{\bfz:g(\bfz;\bftheta) = \bfx \}} p(\bfz|\bftheta) \mu(d\bfz),
\end{equation}
where $\mu$ is a probability measure.
%Importantly, the simulator $g$ is assumed to be a non-invertible function, that can only be
%used to generate data in forward mode. For this reason, evaluating the integral
%in Eqn.~\ref{eqn:p_x_sim} is intractable. As commonly found
%in science, we finally assume the lack of access to or existence of derivatives of $g$ with respect to $\bftheta$,
%e.g. as when $g$ is specified as a computer program.

@@

% In this simple form, stochastic gradient descent based on
% Eqn~\ref{eqn:approx-grad} is known to suffer from premature convergence and lack
% of scale invariance. To avoid the dependence on the parameterization of the
% proposal distribution, natural gradients~\citep{amari1998natural} can be used
% instead to point towards the steepest descent in the space of realizable
% proposal distributions, rather than in the space of their parameters. Effectively,
% the natural evolution strategy (NES) algorithm~\citep{2011arXiv1106.4487W}
% corrects the gradient of $U$ according to the local curvature of the KL-divergence
% surface of proposal distributions, which results in forming
% \begin{equation}\label{eqn:approx-grad-nat}
%     \tilde{\nabla}_\bfpsi U(\bfpsi) = \mathbf{F}(\bfpsi)^{-1} \nabla_\bfpsi U(\bfpsi),
% \end{equation}
% where $\mathbf{F}(\bfpsi)$ is the Fisher information matrix
% \begin{equation}
%     \mathbf{F}(\bfpsi) = \mathbb{E}_{\bftheta \sim q_\bfphi}[ \nabla_\bfpsi \log q_\bfpsi(\bftheta) \nabla_\bfpsi \log q_\bfpsi(\bftheta)^{\top} ]
% \end{equation}
% and can be estimated from samples, in particular by reusing the derivatives $\nabla_\bfpsi \log q_\bfpsi(\bftheta)$ already
% evaluated to form $\nabla_\bfpsi U(\bfpsi)$.

@@

%Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
%have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
%a distribution parameterized in terms of $\bfpsi$:
%\begin{equation}\label{eqn:p_psi}
%    \mathbf{x} \sim \qxpsi \equiv \bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta).
%\end{equation}
%Intuitively, this corresponds to a family of simulators, each configured
%with randomly sampled parameters $\bftheta \sim q(\bftheta|\bfpsi)$, whose joint collection
%of generated samples is optimized with adversarial training to approach the real data distribution $p_r(\mathbf{x})$.
%More formally, the learned model  therefore corresponds to the marginal distribution
%of the generated data. That is,
%\begin{equation}
%    \qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
%\end{equation}
%Intuitively, this corresponds to a family of simulators, each configured
%with randomly sampled parameters $\bftheta \sim q(\bftheta|\bfpsi)$, whose joint collection
%of generated samples is optimized with adversarial training to approach the real data distribution $p_r(\mathbf{x})$.

@@

The variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
the parameterized marginal distribution of the generated data
\begin{equation}
    \qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
\end{equation}
We can think of $q(\bfx|\bfpsi)$ as a \textit{variational program} as described
in \citep{2016arXiv161009033R}, though more complicated than a simple
reparameterization of normally distributed noise $\bfz$ through a differentiable
function. In our case, the variational program is a
marginalized, non-differentiable  simulator.  Its density is  intractable;
nevertheless, it can generate samples for $\bfx$ whose expectations are differentiable with
respect to $\bfpsi$.
Operationally, we sample from this marginal model via
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim \qxpsi \equiv \bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta).
\end{equation}

@@

The observed data is sampled from a Poisson with mean $\lambda^* = 7$.
Algorithm~\ref{alg:avo} is run for 300 epochs with mini-batches of size $M=64$
and the following configuration. For the critic $d$, we use a 3-layer MLP with 10
hidden nodes per layer and ReLU activations. At each epoch, Adam is run for
$n_\text{critic}=100$ iterations with a step size $\alpha=0.01$, decay rates
$\beta_1=\beta_2=0.5$ and its inner first and second moment vectors reset at
each outer epoch in order to avoid building momentum in staled directions.  For
estimating $\lambda^*$, we parameterize $\bftheta$ as $\log(\lambda)$ and use a univariate Gaussian proposal distribution
$q(\bftheta|\bfpsi)$ initialized with a mean at $\log(5)$ and unit variance. At
each epoch, parameters $\bfpsi$ are updated by taking one Adam step, with
$\alpha=0.01$ and $\beta_1=\beta_2=0.5$. The gradient penalty coefficient is set to
$\lambda=0.025$, and the entropy penalty is evaluated at both $\gamma=0$ and $\gamma=5$.

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{figures/poisson.pdf}
    \caption{Discrete Poisson model with unknown mean.
             ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after adversarial variational optimization. For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
                        the true value $\log(\lambda^*)$. Penalizing the entropy of the proposal distribution ($\gamma=5$) results in a tighter density.
             ({\it Top right}) Model distributions $\qxpsi$ after training. This plot shows that the resulting parameterizations of the simulator closely reproduce the true distribution.
             ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
             }\label{fig:poisson}
\end{figure}

@@

Lastly, we make a connection to Operator Variational Inference (OPVI)~\citep{2016arXiv161009033R},
which is a generalization of variational inference formulated as the following optimization problem
\begin{eqnarray}
\bflambda^* = \inf_\bflambda \sup_\bfphi \mathbb{E}_{ \bfz \sim q( \bfz | \bflambda)} [ (O^{p,q} f_\bfphi) ] \, .
\end{eqnarray}
In traditional VI with the KL distance, this corresponds to $(O^{p,q} f) = \log q(\bfz) - \log p(\bfz|\bfx) \, \forall f \in \mathcal{F}$.
AVO can be cast into a similar form with expectations over $\bfx$ instead of $\bfz$ and
\begin{eqnarray} \label{eq:opvi_avo}
\bfpsi^* &=& \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ \bfx \sim q( \bfx | \bfpsi)} [ (O^{p_r,p_\bfpsi} d_\bfphi) ] \\\
\label{eq:opvi_avo2} &=&  \inf_\bfpsi \sup_\bfphi\mathbb{E}_{\bfx\sim p_r(\bfx)}[d(\bfx)] - \mathbb{E}_{\bfx \sim q(\bfx|\bfpsi)}[d(\bfx)] \, .
\end{eqnarray}
Rewriting Eqn.~\ref{eq:opvi_avo} as Eqn.~\ref{eq:opvi_avo2} is possible through importance sampling,
corresponding to an implicit form of the operator
\begin{equation}
(O^{p,q} d) = \left( \frac{p_r(\bfx)}{q(\bfx|\bfpsi)}  -1 \right) d(\bfx) \, ,
\end{equation}
which reinforces the link to density ratio estimation and inference with implicit models.

>>>>

We assume that the stochastic generative process that defines $p(\mathbf{x}|\bftheta)$ is
specified through a non-differentiable deterministic function $g(\cdot; \bftheta) : {\cal Z} \to
\mathbb{R}^d$. Operationally, %That is, we consider
\begin{equation}\label{eqn:p_theta}
    \mathbf{x} \sim p(\mathbf{x}|\bftheta) \equiv \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta)
\end{equation}
such that the density $p(\mathbf{x}|\bftheta)$ can be
written as
%considered as a non-differentiable reparametrization of the random variable $\bfz$
%\begin{equation}\label{eqn:p_x_sim}
%    p(\mathbf{x}|\bftheta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\bfz:g_i(\bfz;\bftheta) \leq x_i\}} p(\bfz|\bftheta) \mu(d\bfz),
%\end{equation}
\begin{equation}\label{eqn:p_x_sim}
    p(\mathbf{x}|\bftheta) = \int_{\{\bfz:g(\bfz;\bftheta) = \bfx \}} p(\bfz|\bftheta) \mu(d\bfz),
\end{equation}
where $\mu$ is a probability measure.
%Importantly, the simulator $g$ is assumed to be a non-invertible function, that can only be
%used to generate data in forward mode. For this reason, evaluating the integral
%in Eqn.~\ref{eqn:p_x_sim} is intractable. As commonly found
%in science, we finally assume the lack of access to or existence of derivatives of $g$ with respect to $\bftheta$,
%e.g. as when $g$ is specified as a computer program.

@@

% In this simple form, stochastic gradient descent based on
% Eqn~\ref{eqn:approx-grad} is known to suffer from premature convergence and lack
% of scale invariance. To avoid the dependence on the parametrization of the
% proposal distribution, natural gradients~\citep{amari1998natural} can be used
% instead to point towards the steepest descent in the space of realizable
% proposal distributions, rather than in the space of their parameters. Effectively,
% the natural evolution strategy (NES) algorithm~\citep{2011arXiv1106.4487W}
% corrects the gradient of $U$ according to the local curvature of the KL-divergence
% surface of proposal distributions, which results in forming
% \begin{equation}\label{eqn:approx-grad-nat}
%     \tilde{\nabla}_\bfpsi U(\bfpsi) = \mathbf{F}(\bfpsi)^{-1} \nabla_\bfpsi U(\bfpsi),
% \end{equation}
% where $\mathbf{F}(\bfpsi)$ is the Fisher information matrix
% \begin{equation}
%     \mathbf{F}(\bfpsi) = \mathbb{E}_{\bftheta \sim q_\bfphi}[ \nabla_\bfpsi \log q_\bfpsi(\bftheta) \nabla_\bfpsi \log q_\bfpsi(\bftheta)^{\top} ]
% \end{equation}
% and can be estimated from samples, in particular by reusing the derivatives $\nabla_\bfpsi \log q_\bfpsi(\bftheta)$ already
% evaluated to form $\nabla_\bfpsi U(\bfpsi)$.

@@

%Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
%have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
%a distribution parametrized in terms of $\bfpsi$:
%\begin{equation}\label{eqn:p_psi}
%    \mathbf{x} \sim \qxpsi \equiv \bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta).
%\end{equation}
%Intuitively, this corresponds to a family of simulators, each configured
%with randomly sampled parameters $\bftheta \sim q(\bftheta|\bfpsi)$, whose joint collection
%of generated samples is optimized with adversarial training to approach the real data distribution $p_r(\mathbf{x})$.
%More formally, the learned model  therefore corresponds to the marginal distribution
%of the generated data. That is,
%\begin{equation}
%    \qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
%\end{equation}
%Intuitively, this corresponds to a family of simulators, each configured
%with randomly sampled parameters $\bftheta \sim q(\bftheta|\bfpsi)$, whose joint collection
%of generated samples is optimized with adversarial training to approach the real data distribution $p_r(\mathbf{x})$.

@@

The variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
the parametrized marginal distribution of the generated data
\begin{equation}
    \qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
\end{equation}
We can think of $q(\bfx|\bfpsi)$ as a \textit{variational program} as described
in \citep{2016arXiv161009033R}, though more complicated than a simple
reparametrization of normally distributed noise $\bfz$ through a differentiable
function. In our case, the variational program is a
marginalized, non-differentiable  simulator.  Its density is  intractable;
nevertheless, it can generate samples for $\bfx$ whose expectations are differentiable with
respect to $\bfpsi$.
Operationally, we sample from this marginal model via
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim \qxpsi \equiv \bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta).
\end{equation}

@@

The observed data is sampled from a Poisson with mean $\lambda^* = 7$.
Algorithm~\ref{alg:avo} is run for 300 epochs with mini-batches of size $M=64$
and the following configuration. For the critic $d$, we use a 3-layer MLP with 10
hidden nodes per layer and ReLU activations. At each epoch, Adam is run for
$n_\text{critic}=100$ iterations with a step size $\alpha=0.01$, decay rates
$\beta_1=\beta_2=0.5$ and its inner first and second moment vectors reset at
each outer epoch in order to avoid building momentum in staled directions.  For
estimating $\lambda^*$, we parametrize $\bftheta$ as $\log(\lambda)$ and use a univariate Gaussian proposal distribution
$q(\bftheta|\bfpsi)$ initialized with a mean at $\log(5)$ and unit variance. At
each epoch, parameters $\bfpsi$ are updated by taking one Adam step, with
$\alpha=0.01$ and $\beta_1=\beta_2=0.5$. The gradient penalty coefficient is set to
$\lambda=0.025$, and the entropy penalty is evaluated at both $\gamma=0$ and $\gamma=5$.

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{figures/poisson.pdf}
    \caption{Discrete Poisson model with unknown mean.
             ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after adversarial variational optimization. For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
                        the true value $\log(\lambda^*)$. Penalizing the entropy of the proposal distribution ($\gamma=5$) results in a tighter density.
             ({\it Top right}) Model distributions $\qxpsi$ after training. This plot shows that the resulting parametrizations of the simulator closely reproduce the true distribution.
             ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
             }\label{fig:poisson}
\end{figure}

@@

Lastly, we make a connection to Operator Variational Inference (OPVI)~\citep{2016arXiv161009033R},
which is a generalization of variational inference formulated as the following optimization problem
\begin{eqnarray}
\bflambda^* = \inf_\bflambda \sup_\bfphi \mathbb{E}_{ \bfz \sim q( \bfz | \bflambda)} [ (O^{p,q} f_\bfphi) ] \, .
\end{eqnarray}
In traditional VI with the KL distance, this corresponds to $(O^{p,q} f) = \log q(\bfz) - \log p(\bfz|\bfx) \, \forall f \in \mathcal{F}$.
AVO can be cast into a similar form with expectations over $\bfx$ instead of $\bfz$ and
\begin{eqnarray} \label{eq:opvi_avo}
\bfpsi^* &=& \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ \bfx \sim q( \bfx | \bfpsi)} [ (O^{p_r,p_\bfpsi} d_\bfphi) ] \nonumber \\
\label{eq:opvi_avo2} &=&  \inf_\bfpsi \sup_\bfphi\mathbb{E}_{\bfx\sim p_r(\bfx)}[d(\bfx)] - \mathbb{E}_{\bfx \sim q(\bfx|\bfpsi)}[d(\bfx)] \, .
\end{eqnarray}
Rewriting Eqn.~\ref{eq:opvi_avo} as Eqn.~\ref{eq:opvi_avo2} is possible through importance sampling,
corresponding to an implicit form of the operator
\begin{equation}
(O^{p,q} d) = \left( \frac{p_r(\bfx)}{q(\bfx|\bfpsi)}  -1 \right) d(\bfx) \, ,
\end{equation}
which reinforces the link to density ratio estimation and inference with implicit models.
